Hydra battles in Coq

### 2020-10-22 doc/hydras.tex
redaction of chapter on Gamma0 (partial)

In~\cite{grimm:hal-00911710}, José Grimm establishes the consistency between or ordinal notations (\texttt{T1} and \texttt{T2} (Veblen normal form) and his implementation
of ordinal numbers after Bourbaki's set theory.

@@

\textbf{To write!}

@@

>>>>

In~\cite{grimm:hal-00911710}, José Grimm establishes the consistency between our ordinal notations (\texttt{T1} and \texttt{T2} (Veblen normal form) and his implementation
of ordinal numbers after Bourbaki's set theory.

@@

\section{Introduction}
We present a notation system for the ordinal $\Gamma_0$, following Chapter V, Section 14 of~\cite{schutte}: ``A notation system fir the ordinals $<\Gamma_0$''.
We try to be as close as possible to Schütte's text and usual practices of \coq{} developments.

The ordinal $\Gamma_0$ is defined in Section 13 of ~\cite{schutte} as the least \emph{strongly critical ordinal}. It is widely known as the \emph{Feferman-Schütte ordinal}.
Section V, 13 of~\cite{schutte} defines \emph{strongly critical} and
\emph{maximal $\alpha$-critical} ordinals.

\noindent\emph{From \href{../src/html/hydras.Schutte.Critical.html\#strongly_critical}%
{\texttt{Schutte.Critical}}}

\begin{Coqsrc}
Definition strongly_critical alpha := In (Cr alpha) alpha.

Definition maximal_critical alpha : Ensemble Ord :=
  fun gamma =>
    In (Cr alpha) gamma /\
    forall ksi, alpha < ksi -> ~ In (Cr alpha) ksi.
\end{Coqsrc}

\begin{project}
In order to define $\Gamma_0$ as a n inhabitant of type \texttt{Ord}, you should prove that the set of strongly critical ordinal is not empty, hence has a least element.
\end{project}

In the present version of this development, we will only study $\Gamma_0$ as a notation system, much more powerful than the ordinal notation for $\epsilon_0$.

%\index{Projects}
%
%\begin{project}
% Schûtte's section 13 of~\cite{schutte} contain several definitions and lemmas which give another view on $\Gamma_0$.  We leave it as a project to implement them in \coq{}.  
%\end{project}

@@

\section{The type \texttt{T2} of ordinal terms}

The notation system for ordinals less than $\gamma_0$ comes from the following theorem of~\cite{schutte}, where $\psi\,\alpha$ is the ordering function 
of the set of maximal $\alpha$-critical ordinals.

\index{Projects}
\begin{project}
This theorem is not formally proved in this development yet. It should be!
\end{project}

\begin{quote}
  Any ordinal $\not= 0$ which is not strongly critical can be expressed in terms of $+$ and $\psi$.
\end{quote}

Like in Chapter~\ref{chap:T1}, we define an inductive type with two constructors, one for $0$, the other for the construction $\psi(\alpha,\beta)\times(n+1)+\gamma$, adapting a Manolios-Vroon-like notation~\cite{Manolios2005} to
\emph{Veblen normal forms}.

\noindent\emph{From \href{../src/html/hydras.Gamma0.T2.html\#T2}%
{\texttt{Gamma0.T2}}}
\begin{Coqsrc}
(**  [gcons alpha beta n gamma] is : [psi(alpha,beta)*(S n)+ gamma]  *)

Inductive T2 : Set :=
| zero : T2
| gcons : T2 -> T2  -> nat -> T2 -> T2.

Notation "[ alpha , beta ]" := (gcons alpha beta 0 zero)
                                 (at level 0): t2_scope.
\end{Coqsrc}

Like in chapter~\ref{chap:T1}, we get familiar with the type \texttt{T2} by recognising simple constructs like finite ordinals, $\omega$, etc., as inhabitants of ÷texttt{T2}.

\begin{Coqsrc}
Notation  "'one'"  := [zero,zero] : T2_scope.

Notation "'FS' n" := (gcons zero zero n zero) (at level 10) : T2_scope.

(** the [n]-th ordinal  *)

Definition fin (n:nat) := match n with 0 => zero | S p => FS p end.

Notation "'omega'"  := [zero,one] : T2_scope.
\end{Coqsrc}

We are also able to give a term to $\epsilon_0$, and even to generalize it!.

\begin{Coqsrc}
Definition epsilon0  := [one,zero].

Definition epsilon alpha := [one, alpha].
\end{Coqsrc}

The justification of the definition of \texttt{epsilon0} will be given in further sections.

\section{How big is \texorpdfstring{$\Gamma_0$}{\texttt{Gamma0}}?}

Let us define a strict order on type \texttt{T2}. The following definition is 
an adaptation of Schütte's, taking into account the multiplications by a natural number (inspired by~\cite{Manolios2005}, and also present in \texttt{T1}).

\begin{Coqsrc}
Inductive lt : T2 -> T2 -> Prop :=
| (* 1 *) 
 lt_1 : forall alpha beta n gamma,  zero t2< gcons alpha beta n gamma
| (* 2 *)
 lt_2 : forall alpha1 alpha2 beta1 beta2 n1 n2 gamma1 gamma2, 
                alpha1 t2< alpha2 ->
                beta1 t2< gcons alpha2 beta2 0 zero ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha2 beta2 n2 gamma2
| (* 3 *)
 lt_3 : forall alpha1  beta1 beta2 n1 n2 gamma1 gamma2, 
               beta1 t2< beta2 ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha1 beta2 n2 gamma2

| (* 4 *)
 lt_4 : forall alpha1 alpha2 beta1 beta2 n1 n2 gamma1 gamma2, 
               alpha2 t2< alpha1 ->
               gcons alpha1 beta1 0 zero t2< beta2 ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha2 beta2 n2 gamma2

| (* 5 *)
lt_5 : forall alpha1 alpha2 beta1 n1 n2 gamma1 gamma2, 
               alpha2 t2< alpha1 ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha2  (gcons alpha1 beta1 0 zero) n2 gamma2

| (* 6 *)
lt_6 : forall alpha1 beta1  n1  n2 gamma1 gamma2,  (n1 < n2)%nat ->
                                    gcons alpha1 beta1 n1 gamma1 t2< 
                                    gcons alpha1 beta1 n2 gamma2

| (* 7 *)
  lt_7 : forall alpha1 beta1 n1   gamma1 gamma2,  gamma1 t2< gamma2 ->
                                      gcons alpha1 beta1 n1 gamma1 t2<
                                      gcons alpha1 beta1 n1 gamma2
where  "o1 t2< o2" := (lt o1 o2): T2_scope.
\end{Coqsrc}

Here are two simple verifications: $\omega$ is less than $\epsilon_0$, whic is less than $\psi(2,1)$.

\begin{Coqsrc}
Goal omega t2< epsilon0.
Proof.
  unfold epsilon0; auto with T2.
Qed.

Goal epsilon0 t2< [2,1].
Proof.
  unfold epsilon0;  apply lt_2.
  - apply lt_6; auto with arith.
  - auto with T2.
Qed.
\end{Coqsrc}

\index{Projects}
\begin{project}
Write a tactic that solves automatically goals of the form (\texttt{$\alpha$ t2< $\beta$}), where $\alpha$ and $\beta$ are closed terms of type \texttt{T2}.
\end{project}
